Retrofitting Effect Handlers onto OCaml

### 2020-10-27 retro-concurrency.tex
Updates

What makes effect handlers quite powerful is that they provide a structured
form of delimited continuations~\cite{}, which enables advanced non-local
control-flow mechanisms such as resumable exceptions, coroutines, generators
and asynchronous I/O to be composably expressed. As an example, consider the
task of handling division-by-zero case in a calculator application. Different
calculators handle division-by-zero differently. In the iOS 13 Calculator App
(|iOS|), division-by-zero returns |Error|. In the Google search calculator App
(|gsearch|) |0/0| returns to |Error|, and for any other |n| not equal to |0|,
|n/0| returns |Infinity| with the same sign as |n|. In the OCaml top-level,
division-by-zero raises |Division_by_zero| exception. How can we capture these
different behaviours without changing the evaluator?

@@

If the evaluator returns a value, we simply return that. In case the evaluator
performs the |DivBy0| effect, we catch the effect in the effect case along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. In the case of |iOS|, we |continue| the suspended computation with
|Error| value. Hence, the computation |ios((1/0)/2) | returns |Error|. On the
other hand, we can get |gsearch| semantics by enclosing the evaluator in the
following handler:

@@

The computation |gsearch((1/0)/2)| returns positive infinity (|Inf P|). We can
extend the calculator modularly with the ability to read user input using the
|Read| effect defined earlier as follows:

@@

\subsection{Retrofitting effect handlers}

@@

The need to host millions of active continuations in the heap rules out the use
of large contiguous stack space for continuations as in C. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate stack. Additionally, we must
generate DWARF stack unwind tables for stacks that may be non-contiguous and
contain frames from different languages. In this work, we develop a series of
techniques to implement efficient effect handlers for OCaml with the help of
compiler and runtime support that satisfy the requirements above.

The search for an expressive effect system which guarantees that all the
effects performed in the program are handled (\emph{effect safety}) in the
presence of advanced features such as polymorphism, modularity and generativity
is an active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety.
For example, running |perform (Read stdin)| without an enclosing handler raises
exception |Unhandled|. We leave the question of effect safety for future work.

@@

%\begin{itemize}
%	\item the design and implementation of affine effect handlers for OCaml with
%		negligible performance impact on code that does not use effect handlers.
%	\item

\section{Background: OCaml Stacks}

\section{Handlers by example}

\section{Semantics}

@@

\section{Implementation}

\section{Evaluation}

\section{Related Work}

\section{Conclusions and Future Work}

>>>>

What makes effect handlers quite powerful is that they provide a structured
form of delimited continuations~\cite{}, which enables advanced non-local
control-flow mechanisms such as resumable exceptions, coroutines, generators
and asynchronous I/O to be composably expressed. As an example, consider the
task of handling division-by-zero case in a calculator application. Different
calculators handle division-by-zero differently. In the iOS 13 Calculator App
(|iOS|), division-by-zero returns |Error|. In the Google search calculator App
(|gsearch|) |0/0| returns to |Error|, and for any other |n| not equal to |0|,
|n/0| returns |Infinity| with the same sign as |n|. How can we capture these
different behaviours without changing the evaluator?

@@

If the evaluator returns a value, we simply return that. In case the evaluator
performs the |DivBy0| effect, we catch the effect in the effect case along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. In the case of |iOS|, we |continue| the suspended computation with
|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
other hand, we can get |gsearch| semantics by enclosing the evaluator in the
following handler:

@@

The computation |gsearch(1/0)| returns positive infinity (|Inf P|). We can
extend the calculator modularly with the ability to read user input using the
|Read| effect defined earlier as follows:

@@

\subsection{Retrofitting effect handlers}
\label{sec:req}

@@

The need to host millions of active continuations in the heap rules out the use
of large contiguous stack space for continuations as in C. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate stack. Additionally, we must
generate DWARF stack unwind tables for stacks that may be non-contiguous and
contain frames from different languages. In this work, we develop the compiler
and runtime support for implementing efficient effect handlers for OCaml.

Our work is also timely. WebAssembly~\ref{} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
We believe that our design choices will inform similar choices to be made in
the other languages.

The search for an expressive effect system that guarantees that all the effects
performed in the program are handled (\emph{effect safety}) in the presence of
advanced features such as polymorphism, modularity and generativity is an
active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety.
For example, running |perform (Read stdin)| without an enclosing handler raises
exception |Unhandled|. We leave the question of effect safety for future work.

@@

\begin{itemize}
	\item the design and implementation of affine effect handlers for OCaml with
		negligible performance impact on code that does not use effect handlers.
		Our design lets us retain OCaml's compatibility with program analysis tools
		that inspect the stack.
	\item a formal operation semantics for the effect handler implementation in
		OCaml.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent and parallel programming.
\end{itemize}

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation ``Multicoe OCaml'' to
distinguish it from stock OCaml. Multicore OCaml uses effect handlers as the
mechanism to express user-level concurrency.

The rest of the paper is organized as follows. We describe the details of the
OCaml program stack in the next section, and highlight the features that the
effect handler implementation must retain in order to provide backwards
compatibility. In \S\ref{sec:handlers}, we motivate the design choice that we
have made for effect handlers in Multicore OCaml with the help of examples.
\S\ref{sec:semantics} presents the formal operational semantics for Multicore
OCaml effect handlers. \S\ref{sec:impl} discusses the compiler and the runtime
system support for implementing effect handlers. \S\ref{sec:eval} presents
extensive performance evaluation of effect handlers against the goals set in
\S\ref{sec:req}. \S\ref{sec:related} and \S\ref{sec:conc} present the related
work and the conclusions.

\section{Background: OCaml Stacks}
\label{sec:stack}

\section{Handlers by example}
\label{sec:handlers}

\section{Semantics}
\label{sec:semantics}

@@

\section{Implementation}
\label{sec:impl}

\section{Evaluation}
\label{sec:eval}

\section{Related Work}
\label{sec:related}

\section{Conclusions and Future Work}
\label{sec:conc}
